miniKanren with fair search strategies

### 2019-05-02 Paper/paper.tex
iDFS->DFSi, because iDFS is already a thing

The syntax of a programming language should reflect its semantics. When writing 
a \conde expression in miniKanren, a programmer would expect all clauses share 
the same chance of being explored, as these clauses are written in 
parallel. The existing search strategy, interleaving DFS (iDFS), however, 
prioritize its clauses by the order how they are written down. Similarly, when 
a \conde is followed by another goal conjunctively, a programmer would expect 
states in parallel share the same chance of being explored. Again, the answers 
by iDFS is different from the expectation. We have devised three new search 
strategies that have different level of fairness in \disj and \conj.

@@

We propose three new search strategies, balanced interleaving DFS (biDFS), fair 
DFS (fDFS), and BFS. They have different characteristics of fairness 
(Table.~\ref{fairness}). We prove that the pure subset of our BFS and the BFS 
proposed by Seres et
al~\citep{seres1999algebra} produce the same result when queried. But our code 
is shorter and runs faster. Because the two BFSs are equivalent, we do not 
distinguish them in most places.

\begin{table}
	\begin{tabular}{|c|c|c|c|c|}
		\hline 
		fairness & iDFS & biDFS & fDFS & BFS \\ 
		\hline 
		disj & unfair & almost-fair & fair & fair \\ 
		\hline 
		conj & unfair & unfair & unfair & fair \\ 
		\hline 
	\end{tabular} 
	\caption{fairness of search strategies}
	\label{fairness}
\end{table}

% Interleaving DFS, the search strategy of \conde in
% miniKanren~\citep{friedman_reasoned_2018}, allocates computational resource
% by the order in which the clauses are written. Each clause of a
% \conde takes half of the current resource and passes the other half
% to its following clauses, except for the last clause that takes all of the 
%current resource. 
% The biased treatment provides both opportunity and burden: miniKanren users 
%can place more frequently used 
% goal at the beginning to optimize their programs; however, it might be a 
% catastrophe if a goal that generates many useless states is placed before 
%more 
% important goals. Seasoned miniKanreners usually know how to utilize 
% the unfairness to optimize their programs. However, we believe search 
%strategies that is less sensitive to goal order can also be useful to little 
%miniKanreners as well as seasoned ones. We propose two such search strategies, 
%balanced interleaving DFS (biDFS) and breadth-first search (BFS), and observe 
%how they affect the efficiency and the answer order of known miniKanren 
%programs. The experiment is conducted with the miniKanren from \textit{The 
%Reasoned Schemer, 2nd Edition}.

@@

In the following program, the three \conde clauses differ in a trivial way. So 
we expect lists of each sort constitute $1/4$ of the answer list. However, 
iDFS, the current search strategy, gives us much more lists of \texttt{a} than 
other sorts of lists. And some sorts (e.g. lists of \texttt{c}) are hardly 
found. The situation would be worse if we add more \conde clauses.

@@

We borrow the definition of \emph{fair} \disj{} from \cite{seres1999algebra}: 
search strategies with \emph{fair} \disj{} should allocate resource evenly 
among disjunctive goals. Running the same program with fDFS and BFS give the 
following result.

@@

We conceive a middle place between fair and unfair -- search strategies with 
\emph{almost-fair} \disj{} should allocate resource so evenly among disjunctive 
goals\citep{seres1999algebra} that the the maximal ratio of resource is bounded 
by a constant. Our new search strategy, biDFS, has almost-fair \disj{}. It is 
fair when the number of goals is a power of 2, otherwise, some goals are 
allocated twice more resource than the others. In the previous example, it 
gives the same result. And in the following example, where the \conde has 5 
clause, the clauses of \texttt{b}, \texttt{c}, and \texttt{d} are allocated 
more resource.

@@

In the following program, the three \conde clauses differ in a trivial way. So 
we expect lists of each sort constitute $1/4$ of the answer list. However, 
search strategies with unfair \conj{} (e.g. iDFS, fDFS) give us much 
more lists of \texttt{a} than other sorts of lists. And some sorts (e.g. lists 
of \texttt{c}) are hardly found. Although iDFS's \disj{} is unfair in general, 
it is fair when there is no call to relational definition in sub-goals, 
including this case. The situation would be worse if we add more \conde 
clauses. The result with biDFS, whose \conj{} is also unfair, is similar, but 
due to its different \disj{}, the position of \texttt{b} and \texttt{c} are 
swapped.

@@

\begin{figure}
	\lstinputlisting{Figures/balanced-disj.rkt}
	\caption{biDFS implementation}
	\label{balanced-disj}
\end{figure}

Balanced interleaving DFS (biDFS) has almost-fair \disj{} and unfair \conj{}. 
The implementation of biDFS differs from iDFS in the \disj{} macro. We list the 
new \disj{} with its helpers in Fig.~\ref{balanced-disj}. The first helper 
function, \texttt{split}, takes a list of goals \texttt{ls} and a procedure 
\texttt{k}, partitions \texttt{ls} into two sub-lists of roughly equal length, 
and returns the application of \texttt{k} to the two sub-lists. \texttt{disj*} 
takes a non-empty list of goals \texttt{gs} and returns a goal. With the help 
of \texttt{split}, it essentially constructs a \emph{balanced} binary tree 
where leaves are elements of \texttt{gs} and nodes are \texttt{disj2}, whence 
the name of this search strategy. In contrast, the \disj{} in iDFS essentially 
constructs the same sort of binary tree in one of the most unbalanced forms.

@@

\begin{figure}
	\lstinputlisting{Figures/fDFS.rkt}
	\caption{fDFS implementation}
	\label{fDFS}
\end{figure}

Fair DFS (fDFS) has fair \disj{} and unfair \conj{}. The implementation of fDFS 
differs from iDFS's in \texttt{disj2} (Fig.~\ref{fDFS}). \texttt{disj2} is 
changed to call a new and fair version of \texttt{append-inf}. 
\texttt{append-inf/fair} immediately calls its helper, 
\texttt{append-inf/fair\^{}}, with the first argument, \texttt{s?}, set to 
\texttt{\#{}t}, which indicates that \texttt{s-inf} and \texttt{t-inf} haven't 
been swapped. The swapping happens at the third \texttt{cond} clause in the 
helper, where \texttt{s?} is updated accordingly. The first two \texttt{cond} 
clauses essentially copy the \texttt{car}s and stop recursion when one of the 
input spaces is obviously finite. The third clause, as we mentioned early, is 
just for swapping. When the fourth and last clause runs, we know that both 
\texttt{s-inf} and \texttt{t-inf} are ended with a thunk. In this case, a new 
thunk is constructed. The new thunk calls the driver recursively. Here changing 
the order of \texttt{t-inf} and \texttt{s-inf} won't hurt the fairness (though 
it will change the order of answers). We swapped them back so that answers are 
produced in a more natural order.

@@

BFS is fair in both \disj{} and \conj{}. Our implementation is based on 
fDFS (not iDFS). All we have to do is apply two trivial changes to 
\texttt{append-map-inf}. First, rename it to \texttt{append-map-inf/fair}. 
Second, replace its use of \texttt{append-inf} to \texttt{append-inf/fair}.

@@

\begin{table}
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline 
		benchmark & size & iDFS & biDFS & fDFS & optimized BFS & Silvija's BFS  
		\\
		\hline
		very-recursiveo & 100000 &  579 &  793 & 2131 & 1438 & 3617 \\
		& 200000 & 1283 & 1610 & 3602 & 2803 & 4212 \\
		& 300000 & 2160 & 2836 &    - & 6137 &    - \\
		\hline 
		appendo  & 100 &  31 &  41 &  42 &  31 &  68 \\ 
		& 200 & 224 & 222 & 221 & 226 & 218 \\ 
		& 300 & 617 & 634 & 593 & 631 & 622 \\ 
		\hline 
		reverso & 10 &   5 &   3 &   3 &     38 &     85 \\ 
		& 20 & 107 &  98 &  51 &   4862 &   5844 \\
		& 30 & 446 & 442 & 485 & 123288 & 132159 \\ 
		\hline
		quine-1 & 1 &  71 &  44 & 69 & - & - \\ 
		& 2 & 127 & 142 & 95 & - & - \\ 
		& 3 & 114 & 114 & 93 & - & - \\ 
		\hline
		quine-2 & 1 & 147 & 112 &  56 & - & - \\ 
		& 2 & 161 & 123 & 101 & - & - \\ 
		& 3 & 289 & 189 & 104 & - & - \\ 
		\hline 
		'(I love you)-1 &  99 & 56 & 15 & 22 &  74 & 165 \\ 
		& 198 & 53 & 72 & 55 &  47 &  74 \\
		& 297 & 72 & 90 & 44 & 181 & 365 \\ 
		\hline
		'(I love you)-2 &  99 & 242 &  61 & 16 &  66 &  99 \\ 
		& 198 & 445 & 110 & 60 &  42 &  64 \\
		& 297 & 476 & 146 & 49 & 186 & 322 \\ 
		\hline 
	\end{tabular}
	\caption{The results of a quantitative evaluation: running times of 
	benchmarks 
		in milliseconds}
	\label{compare-efficiency}
\end{table}

In this section, we compare the efficiency of search strategies. A concise 
description is in Table~\ref{compare-efficiency}. A hyphen means running out of 
memory. The first two benchmarks are taken from 
\citep{friedman_reasoned_2018}. \texttt{reverso} is from 
\citep{rozplokhas2018improving}. Next two benchmarks 
about quine are modified from a similar test case in \citep{byrd2017unified}. 
The modifications are made 
to circumvent the need for symbolic constraints (e.g. $\neq$, 
\texttt{absento}). Our version generates de 
Bruijnized expressions and prevent closures getting into list. The two 
benchmarks differ in the \conde clause order of their relation interpreters. 
The last two 
benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love 
you)}. This benchmark is also inspired by \citep{byrd2017unified}. Again, the 
sibling benchmarks differ in the \conde clause order of their relation 
interpreters. The first one 
has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the 
end, while the other has them at the beginning. We conjecture that iDFS would 
perform badly in the second case because elimination rules complicate the 
problem when running backward. The evaluation supports our conjecture.

In general, only iDFS and biDFS constantly perform well. fDFS is just as 
efficient in all benchmarks but \texttt{very-recursiveo}. Both BFS have obvious 
overhead in many cases. Among the three variants of DFS (they all have unfair 
\conj{}), fDFS is most resistant to clause permutation, followd by biDFS then 
iDFS. Among the two implementation of BFS, ours constantly performs as well or 
better. Interestingly, every strategies with fair \disj{} suffers in 
\texttt{very-recursiveo} and fDFS performs well elsewhere. Therefore, this 
benchmark might be a special case. Fair \conj{} imposes overhead constantly 
except in \texttt{appendo}. The reason might be that strategies with fair 
\conj{} tend to keep more intermediate states in the memory.

@@

We devise three new search strategies: balanced interleaving DFS (biDFS), fair 
DFS (fDFS), and BFS. biDFS has almost-fair \disj{} and unfair \conj{}. fDFS has 
fair \disj{} and unfair \conj{}. BFS has both fair \disj{} and fair \conj{}.

Our quantitative evaluation shows that biDFS and fDFS are competitive 
alternatives to iDFS, the current search strategy, and that BFS is less 
practical.

>>>>

The syntax of a programming language should reflect its semantics. When writing 
a \conde expression in miniKanren, a programmer would expect all clauses share 
the same chance of being explored, as these clauses are written in 
parallel. The existing search strategy, interleaving DFS (DFS$_\textrm{i}$), however, 
prioritize its clauses by the order how they are written down. Similarly, when 
a \conde is followed by another goal conjunctively, a programmer would expect 
states in parallel share the same chance of being explored. Again, the answers 
by DFS$_\textrm{i}$ is different from the expectation. We have devised three new search 
strategies that have different level of fairness in \disj and \conj.

@@

We propose three new search strategies, balanced interleaving DFS (DFS$_\textrm{bi}$), fair 
DFS (DFS$_\textrm{f}$), and BFS. They have different characteristics of fairness 
(Table.~\ref{fairness}). We prove that the pure subset of our BFS and the BFS 
proposed by Seres et
al~\citep{seres1999algebra} produce the same result when queried. But our code 
is shorter and runs faster. Because the two BFSs are equivalent, we do not 
distinguish them in most places.

\begin{table}
	\begin{tabular}{|c|c|c|c|c|}
		\hline 
		fairness & DFS$_\textrm{i}$ & DFS$_\textrm{bi}$ & DFS$_\textrm{f}$ & BFS \\ 
		\hline 
		disj & unfair & almost-fair & fair & fair \\ 
		\hline 
		conj & unfair & unfair & unfair & fair \\ 
		\hline 
	\end{tabular} 
	\caption{fairness of search strategies}
	\label{fairness}
\end{table}

% Interleaving DFS, the search strategy of \conde in
% miniKanren~\citep{friedman_reasoned_2018}, allocates computational resource
% by the order in which the clauses are written. Each clause of a
% \conde takes half of the current resource and passes the other half
% to its following clauses, except for the last clause that takes all of the 
%current resource. 
% The biased treatment provides both opportunity and burden: miniKanren users 
%can place more frequently used 
% goal at the beginning to optimize their programs; however, it might be a 
% catastrophe if a goal that generates many useless states is placed before 
%more 
% important goals. Seasoned miniKanreners usually know how to utilize 
% the unfairness to optimize their programs. However, we believe search 
%strategies that is less sensitive to goal order can also be useful to little 
%miniKanreners as well as seasoned ones. We propose two such search strategies, 
%balanced interleaving DFS (DFS$_\textrm{bi}$) and breadth-first search (BFS), and observe 
%how they affect the efficiency and the answer order of known miniKanren 
%programs. The experiment is conducted with the miniKanren from \textit{The 
%Reasoned Schemer, 2nd Edition}.

@@

In the following program, the three \conde clauses differ in a trivial way. So 
we expect lists of each sort constitute $1/4$ of the answer list. However, 
DFS$_\textrm{i}$, the current search strategy, gives us much more lists of \texttt{a} than 
other sorts of lists. And some sorts (e.g. lists of \texttt{c}) are hardly 
found. The situation would be worse if we add more \conde clauses.

@@

We borrow the definition of \emph{fair} \disj{} from \cite{seres1999algebra}: 
search strategies with \emph{fair} \disj{} should allocate resource evenly 
among disjunctive goals. Running the same program with DFS$_\textrm{f}$ and BFS give the 
following result.

@@

We conceive a middle place between fair and unfair -- search strategies with 
\emph{almost-fair} \disj{} should allocate resource so evenly among disjunctive 
goals\citep{seres1999algebra} that the the maximal ratio of resource is bounded 
by a constant. Our new search strategy, DFS$_\textrm{bi}$, has almost-fair \disj{}. It is 
fair when the number of goals is a power of 2, otherwise, some goals are 
allocated twice more resource than the others. In the previous example, it 
gives the same result. And in the following example, where the \conde has 5 
clause, the clauses of \texttt{b}, \texttt{c}, and \texttt{d} are allocated 
more resource.

@@

In the following program, the three \conde clauses differ in a trivial way. So 
we expect lists of each sort constitute $1/4$ of the answer list. However, 
search strategies with unfair \conj{} (e.g. DFS$_\textrm{i}$, DFS$_\textrm{f}$) give us much 
more lists of \texttt{a} than other sorts of lists. And some sorts (e.g. lists 
of \texttt{c}) are hardly found. Although DFS$_\textrm{i}$'s \disj{} is unfair in general, 
it is fair when there is no call to relational definition in sub-goals, 
including this case. The situation would be worse if we add more \conde 
clauses. The result with DFS$_\textrm{bi}$, whose \conj{} is also unfair, is similar, but 
due to its different \disj{}, the position of \texttt{b} and \texttt{c} are 
swapped.

@@

\begin{figure}
	\lstinputlisting{Figures/balanced-disj.rkt}
	\caption{DFS$_\textrm{bi}$ implementation}
	\label{balanced-disj}
\end{figure}

Balanced interleaving DFS (DFS$_\textrm{bi}$) has almost-fair \disj{} and unfair \conj{}. 
The implementation of DFS$_\textrm{bi}$ differs from DFS$_\textrm{i}$ in the \disj{} macro. We list the 
new \disj{} with its helpers in Fig.~\ref{balanced-disj}. The first helper 
function, \texttt{split}, takes a list of goals \texttt{ls} and a procedure 
\texttt{k}, partitions \texttt{ls} into two sub-lists of roughly equal length, 
and returns the application of \texttt{k} to the two sub-lists. \texttt{disj*} 
takes a non-empty list of goals \texttt{gs} and returns a goal. With the help 
of \texttt{split}, it essentially constructs a \emph{balanced} binary tree 
where leaves are elements of \texttt{gs} and nodes are \texttt{disj2}, whence 
the name of this search strategy. In contrast, the \disj{} in DFS$_\textrm{i}$ essentially 
constructs the same sort of binary tree in one of the most unbalanced forms.

@@

\begin{figure}
	\lstinputlisting{Figures/fDFS.rkt}
	\caption{DFS$_\textrm{f}$ implementation}
	\label{fDFS}
\end{figure}

Fair DFS (DFS$_\textrm{f}$) has fair \disj{} and unfair \conj{}. The implementation of DFS$_\textrm{f}$ 
differs from DFS$_\textrm{i}$'s in \texttt{disj2} (Fig.~\ref{fDFS}). \texttt{disj2} is 
changed to call a new and fair version of \texttt{append-inf}. 
\texttt{append-inf/fair} immediately calls its helper, 
\texttt{append-inf/fair\^{}}, with the first argument, \texttt{s?}, set to 
\texttt{\#{}t}, which indicates that \texttt{s-inf} and \texttt{t-inf} haven't 
been swapped. The swapping happens at the third \texttt{cond} clause in the 
helper, where \texttt{s?} is updated accordingly. The first two \texttt{cond} 
clauses essentially copy the \texttt{car}s and stop recursion when one of the 
input spaces is obviously finite. The third clause, as we mentioned early, is 
just for swapping. When the fourth and last clause runs, we know that both 
\texttt{s-inf} and \texttt{t-inf} are ended with a thunk. In this case, a new 
thunk is constructed. The new thunk calls the driver recursively. Here changing 
the order of \texttt{t-inf} and \texttt{s-inf} won't hurt the fairness (though 
it will change the order of answers). We swapped them back so that answers are 
produced in a more natural order.

@@

BFS is fair in both \disj{} and \conj{}. Our implementation is based on 
DFS$_\textrm{f}$ (not DFS$_\textrm{i}$). All we have to do is apply two trivial changes to 
\texttt{append-map-inf}. First, rename it to \texttt{append-map-inf/fair}. 
Second, replace its use of \texttt{append-inf} to \texttt{append-inf/fair}.

@@

\begin{table}
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline 
		benchmark & size & DFS$_\textrm{i}$ & DFS$_\textrm{bi}$ & DFS$_\textrm{f}$ & optimized BFS & Silvija's BFS  
		\\
		\hline
		very-recursiveo & 100000 &  579 &  793 & 2131 & 1438 & 3617 \\
		& 200000 & 1283 & 1610 & 3602 & 2803 & 4212 \\
		& 300000 & 2160 & 2836 &    - & 6137 &    - \\
		\hline 
		appendo  & 100 &  31 &  41 &  42 &  31 &  68 \\ 
		& 200 & 224 & 222 & 221 & 226 & 218 \\ 
		& 300 & 617 & 634 & 593 & 631 & 622 \\ 
		\hline 
		reverso & 10 &   5 &   3 &   3 &     38 &     85 \\ 
		& 20 & 107 &  98 &  51 &   4862 &   5844 \\
		& 30 & 446 & 442 & 485 & 123288 & 132159 \\ 
		\hline
		quine-1 & 1 &  71 &  44 & 69 & - & - \\ 
		& 2 & 127 & 142 & 95 & - & - \\ 
		& 3 & 114 & 114 & 93 & - & - \\ 
		\hline
		quine-2 & 1 & 147 & 112 &  56 & - & - \\ 
		& 2 & 161 & 123 & 101 & - & - \\ 
		& 3 & 289 & 189 & 104 & - & - \\ 
		\hline 
		'(I love you)-1 &  99 & 56 & 15 & 22 &  74 & 165 \\ 
		& 198 & 53 & 72 & 55 &  47 &  74 \\
		& 297 & 72 & 90 & 44 & 181 & 365 \\ 
		\hline
		'(I love you)-2 &  99 & 242 &  61 & 16 &  66 &  99 \\ 
		& 198 & 445 & 110 & 60 &  42 &  64 \\
		& 297 & 476 & 146 & 49 & 186 & 322 \\ 
		\hline 
	\end{tabular}
	\caption{The results of a quantitative evaluation: running times of 
	benchmarks 
		in milliseconds}
	\label{compare-efficiency}
\end{table}

In this section, we compare the efficiency of search strategies. A concise 
description is in Table~\ref{compare-efficiency}. A hyphen means running out of 
memory. The first two benchmarks are taken from 
\citep{friedman_reasoned_2018}. \texttt{reverso} is from 
\citep{rozplokhas2018improving}. Next two benchmarks 
about quine are modified from a similar test case in \citep{byrd2017unified}. 
The modifications are made 
to circumvent the need for symbolic constraints (e.g. $\neq$, 
\texttt{absento}). Our version generates de 
Bruijnized expressions and prevent closures getting into list. The two 
benchmarks differ in the \conde clause order of their relation interpreters. 
The last two 
benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love 
you)}. This benchmark is also inspired by \citep{byrd2017unified}. Again, the 
sibling benchmarks differ in the \conde clause order of their relation 
interpreters. The first one 
has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the 
end, while the other has them at the beginning. We conjecture that DFS$_\textrm{i}$ would 
perform badly in the second case because elimination rules complicate the 
problem when running backward. The evaluation supports our conjecture.

In general, only DFS$_\textrm{i}$ and DFS$_\textrm{bi}$ constantly perform well. DFS$_\textrm{f}$ is just as 
efficient in all benchmarks but \texttt{very-recursiveo}. Both BFS have obvious 
overhead in many cases. Among the three variants of DFS (they all have unfair 
\conj{}), DFS$_\textrm{f}$ is most resistant to clause permutation, followd by DFS$_\textrm{bi}$ then 
DFS$_\textrm{i}$. Among the two implementation of BFS, ours constantly performs as well or 
better. Interestingly, every strategies with fair \disj{} suffers in 
\texttt{very-recursiveo} and DFS$_\textrm{f}$ performs well elsewhere. Therefore, this 
benchmark might be a special case. Fair \conj{} imposes overhead constantly 
except in \texttt{appendo}. The reason might be that strategies with fair 
\conj{} tend to keep more intermediate states in the memory.

@@

We devise three new search strategies: balanced interleaving DFS (DFS$_\textrm{bi}$), fair 
DFS (DFS$_\textrm{f}$), and BFS. DFS$_\textrm{bi}$ has almost-fair \disj{} and unfair \conj{}. DFS$_\textrm{f}$ has 
fair \disj{} and unfair \conj{}. BFS has both fair \disj{} and fair \conj{}.

Our quantitative evaluation shows that DFS$_\textrm{bi}$ and DFS$_\textrm{f}$ are competitive 
alternatives to DFS$_\textrm{i}$, the current search strategy, and that BFS is less 
practical.
